---
title: "Tutorial for Enrichmate"
author: "Hojin Lee"
output: html_document
date: "2025-02-03"
---

# Load Packages

```{r message=FALSE}
library(dplyr)
library(plyr)
library(fgsea)
library(tibble)
library(ggplot2)
library(GO.db)
library(GOSemSim)
library(WriteXLS)
library(colorRamp2)
library(simplifyEnrichment)
library(ComplexHeatmap)
```

# Set environment

To conduct Enrichmate analysis, GOBP databases in different formats are necessary.
Through the following code, the environment for analysis is set up.

```{r}
setwd("/Users/hojin/Dropbox/project/enrichmate/20250203/")
source("/Users/hojin/Dropbox/project/enrichmate/20250203/20250203_enrichmate_function_hj.R")

gogmt <- gmtPathways("/Users/hojin/Dropbox/project/enrichmate/20250203/GeneOntology/v2024.1.Hs/GOBP/c5.go.bp.v2024.1.Hs.symbols.gmt")
goid_terms_df <- read.table("/Users/hojin/Dropbox/project/enrichmate/20250203/GeneOntology/v2024.1.Hs/GOBP/20250124_c5.go.bp.v2024.1.Hs.term.id_mapping_hj.txt",header = T, sep = "\t")

df <- readRDS("/Users/hojin/Dropbox/project/enrichmate/20250203/GeneOntology/v2024.1.Hs/GOBP/20250124_c5.go.bp.v2024.1.Hs_ANCESTOR_df_hj.rds")
goid_level_df <- readRDS("/Users/hojin/Dropbox/project/enrichmate/20250203/GeneOntology/v2024.1.Hs/GOBP/20250124_c5.go.bp.v2024.1.Hs_LEVEL_df_hj.rds")

goid_ancestor_level_df <- merge(goid_level_df, df, by.x = "GOID", by.y = "ANCESTOR_GOID", all.y = T)
colnames(goid_ancestor_level_df) <- c("ANCESTOR_GOID", "ANCESTOR_LEVEL", "TARGET_GOID", "ANCESTOR_GOTERM")
goid_ancestor_level_df <- goid_ancestor_level_df %>% dplyr::filter(ANCESTOR_LEVEL != "all")

goid_ancestor_info <- goid_terms_df$GOID[which(goid_terms_df$GOID %in% df$TARGET_GOID)]
goid_terms_df <- goid_terms_df %>% dplyr::filter(GOID %in% goid_ancestor_info)
gogmt <- gogmt[goid_terms_df$GOTERM]

if (!("hsGO" %in% ls())) {
  set.seed(1234); hsGO <- godata('org.Hs.eg.db', ont="BP", computeIC = T) # make IC score using GOsemsim package 
}

```

# Obtain GOBP sets

After conducting GSEA or ORA, significant GO terms were obtained.
The GO IDs (GO:xxxxx) must be in vector format.
The database for this analysis is constructed based on MsigDB data; therefore, only GO IDs included in MsigDB should be retained.
Additionally, ancestor terms are used to obtain representative terms, so GO IDs containing information on ancestor terms must also be retained.

```{r}

tb <- read.csv("/Users/hojin/Dropbox/project/enrichmate/goblet2_adult_0.5_gProfiler_hsapiens_1-23-2025_4-36-35 PM__intersections_final.csv", sep = ",", header = T)

tb %>% dplyr::select(c("source", "term_name", "term_id", "adjusted_p_value")) %>% head()

tb <- tb %>% dplyr::filter(source == "GO:BP")

GO_test <- goid_terms_df %>% dplyr::filter(GOID %in% tb$term_id) %>% dplyr::pull(GOID)

GO_test <- GO_test[GO_test %in% goid_ancestor_level_df$TARGET_GOID] # subset GO terms have ancestor terms

head(GO_test)
```

# Enrichmate analysis with outlier detection

Calculate similarity matrix.
```{r}
# step 1 )
w <- enrichmate_sim_mat(goids = GO_test, hsGO = hsGO)
```

Select the number of clusters.
```{r}
k_val <- 20
```

Distribution of small clusters.
```{r}
# step 2 )
enrichmate_outlier_plot(w = w)
```

Conduct combined clustering.
```{r message=FALSE}
# step 3 )
step3 <- enrichmate_clustering(cutoff = 0.84, w = w, k_vals = k_val, outlier_detect = T, goid_terms_df = goid_terms_df, goid_level_df = goid_level_df)
```

Define representative terms.
```{r message=FALSE}
# step 4 ) 
step4 <- enrichmate_representative_terms(final_res_df = step3$final_res_df,
                                         k_val = k_val,
                                         goid_ancestor_level_df = goid_ancestor_level_df,
                                         outlier_detect = T,
                                         pseudoname_outliers_df = step3$pseudoname_outliers_df,
                                         representative_term_level_cutoff = 2,
                                         GO_explain = 3)
```

Make heatmap and summary files.
```{r fig.width=20, fig.height=10}
# step 5 ) 
step5 <- enrichmate_go_heatmap_summary(w = w,
                                       k_val = k_val,
                                       outlier_detect = T,
                                       ancestor_annotation = T,
                                       filename1 = "total_GOBP_hj.xlsx",
                                       filename2 = "representative_term_hj.xlsx",
                                       plot_pdf = F, # if you want to save plot to pdf, assign T
                                       heatmap_filename = "plot.pdf",
                                       pseudoname_outliers_df = step3$pseudoname_outliers_df,
                                       rep_df = step4,
                                       goid_terms_df = goid_terms_df, font_size = 4)

```

